---
output:
  word_document: default
  html_document: default
---
## Econ 8600 - Homework 3

#### Due EOD 11/19/24

#### {Sarah Grace Turner}


#### **Pascal's double sixes with a weighted die (15 points)**

Modify the code from class to test the case where only one die is fair, while the other die has a 17.5% chance of rolling a 6 and a 16.5% chance for each of the other numbers. Would this be enough to flip the odds and make it a winning wager to bet on getting at least one double six with 24 rolls?

```{r}
set.seed(999)
rollDiefair = function() {
  # 1. Pick a space of possible samples (domain).
  domain = 1:6
  
  # 2. Randomly sample in that space using a probability distribution.
  value = sample(1:6, 1)
  return(value)
}

rollDieBiased = function() {
  domain = 1:6
  probs = c(rep(0.165, 5), 0.175)
  value = sample(1:6, 1, prob=probs)
  return(value)
}

# set up function for Monte Carlo experiment to answer Pascal's double sixes question
doubleSixProb = function(numTrials) {
  numwins = 0
  for (i in 1:numTrials) {
    # 2. Randomly sample in that space using a probability distribution.
    for (j in 1:24) {
      d1 = rollDiefair()
      d2 = rollDieBiased()
      
      # 3. Perform defined operations on the random samples.
      if (d1+d2 == 12) {
        # 4. Aggregate the data.
        numwins = numwins + 1
        break
      }
    }
  }
  # 4. Aggregate the data.
  percent = numwins/numTrials
  return(percent)
}

# run simulation 10,000 times
doubleSixProb(10000)
```
Because the probability is > 50% ( about 51.43%) it's a winning wager to bet on getting at least 1 double six in 24 rolls because you have a slightly higher chance of getting at least one than of not getting one.

#### **Interest Rate Modeling (20 points)**

Suppose an investor wants to simulate the potential future values of an investment with compounding interest. Set up a Monte Carlo simulation where the annual return on the investment is normally distributed with a mean of 6% and a standard deviation of 10%. Simulate the investment's value after 10 years if the initial amount is \$10,000. What is the probability that the investment will be worth more than \$20,000 after 10 years

```{r}
set.seed(123)  
initial_amount <- 10000  
annual_mean <- 0.06      
annual_sd <- 0.10       
years <- 10           
num_simulations <- 100000
threshold <- 20000  #value to exceed


simulate_investment <- function() {
  # gen 10 random ann returns from normal distribution
  annual_returns <- rnorm(years, mean = annual_mean, sd = annual_sd)
  #compounded value after 10 years
  final_value <- initial_amount * prod(1 + annual_returns)
  return(final_value)
}

#run
results <- replicate(num_simulations, simulate_investment())

#prob exceed threshold
probability <- mean(results > threshold)

# display results
print(paste("Probability that the investment exceeds $", threshold, "after", years," years is:", probability))
```

#### **Project Management (35 points total)** Describe distribution in own words. Comment on histogram, don't just plot.

You are tasked with estimating the total duration of a four-phase project where each phase has its own time estimate and dependencies on previous phases. Your task is to use a Monte Carlo simulation to estimate the likely completion time of the project based on the following information:

-   **Phase 1**: Expected to take between 3 and 5 weeks, with an equal chance for any duration in this range (uniform distribution).

-   **Phase 2**:

    -   If Phase 1 finishes in less than 30 days, this phase will take exactly 1 week.

    -   If Phase 1 takes 30 days or more, Phase 2 will take 2 weeks.

-   **Phase 3**: Expected to take around 3 weeks with some variability, modeled as a normal distribution with a mean of 3 weeks and a standard deviation of 2 days.

-   **Phase 4**: This final phase will take between one quarter and one third of the combined time of Phases 1, 2, and 3, with an equal probability within this range.
#Don't just report at end, for every iteration, save value from that, 
Run a Monte Carlo simulation with 1,000 trials to estimate the likely total duration of the project and save the trial results in a vector (length = 1,000, values = estimate from each trial). Answer the following questions:

1.  What is the average completion time across all simulations? (10 points)

  10.84 weeks

2.  What is the probability that the project will take longer than 12 weeks to complete? (10 points)
  
  30.7%

3.  Plot a histogram of the completion times for all 1,000 simulations and describe the distribution in your own words. (15 points)

  The distribution has 2 relative maxima/peaks (looks like a bimodal distribution) around the 10 and the 12.5 week mark, with probability densities that look relatively normally distributed around both, with more falling around the 10-week mode. Either takes 10 or 12 weeks, not around 11 weeks.
  
```{r, Project Management}
set.seed(123)
num_trials <- 1000

# Phase 1: uniform distribution (3-5 weeks)
phase_1_density <- density(runif(10000, min = 3, max = 5))
#runif(1,3,5)

# Phase 3: norm distribution (mean = 3 weeks, sd = 2 days)
phase_3_density <- density(rnorm(10000, mean = 3, sd = 2 / 7))

#sim a single trial
simulate_project_duration <- function() {
  # Phase 1: samp from phase 1 density
  phase_1_duration <- sample(phase_1_density$x, 1, prob = phase_1_density$y)
  
  # Phase 2: depends on phase 1
  if (phase_1_duration < 30 / 7) {  # convert 30 days to weeks
    phase_2_duration <- 1
  } else {
    phase_2_duration <- 2
  }
  
  # Phase 3: samp from phase 3 density
  phase_3_duration <- sample(phase_3_density$x, 1, prob = phase_3_density$y)
  
  # Phase 4: fraction of combined time of all phases
  combined_time <- phase_1_duration + phase_2_duration + phase_3_duration
  phase_4_duration <- runif(1, min = combined_time / 4, max = combined_time / 3)
  
  # Total duration
  total_duration <- phase_1_duration + phase_2_duration + phase_3_duration + phase_4_duration
  return(total_duration)
}

# Run
project_durations <- replicate(num_trials, simulate_project_duration())

# Average completion time
average_completion_time <- mean(project_durations)

# Probability of taking over 12 weeks
probability_over_12 <- mean(project_durations > 12)

# histogram with density overlay
hist(project_durations, breaks = 30, probability = TRUE, col = "skyblue", 
     main = "Histogram of Project Completion Times", xlab = "Completion Time (weeks)")
lines(density(project_durations), col = "red", lwd = 2)
abline(v = average_completion_time, col = "blue", lty = 2, lwd = 2)

#annotations for prettiness
text(average_completion_time, 0.02, paste("Mean =", round(average_completion_time, 2)), col = "blue", pos = 4)
text(15, 0.01, paste("P(T > 12) =", round(probability_over_12, 2)), col = "red", pos = 4)

# output results
print(paste("Average Completion Time:", round(average_completion_time, 2), "weeks"))
print(paste("Probability of Completion > 12 weeks:", round(probability_over_12 * 100, 2), "%"))
```
#phase_4_min = (phase_1_time + phase_2_time + phase_3_time)/4

prob over 12 weeks = mean(total_times > 12)

#### Precipitation (30 points)

Homework3_data.csv, sourced from <https://wrcc.dri.edu/cgi-bin/cliMAIN.pl?sc1770>, contains monthly total precipitation values recorded at Clemson University since 1896. Use this data to write a function to predict precipitation values for each month based on the distribution of historical rainfall for that month (hint: use the density function we used in class). Using your new function, set up a Monte Carlo simulation to find the probability that more than one quarter of the annual rainfall will come from a single month. 

predict_monthly_precip = function(month) {
  dens = density(month, na.rm = T)
  sample(dens$x, size = 1, prob = dens$y)
}

#subset to montly data only 
monthly_data = data[,2:13]
predict_precip = function(){
  

predict_precip = function(){
 monthly_values = apply(monthly_data, 2, predict_monthly_precip)
 return(monthly_values)
}

precip_monte_carlo = function(numTrials)a {
  count = 0 
  
  for (trial in 1:numTrials) {
  precip = predict()
}
}

#note: we're treating all months as independent, but
```{r}
data <- read.csv("C://Users//sgtur//OneDrive - Clemson University//R - Fall 2024//ECON8600//Datasets//Homework3_data.csv", header = TRUE)

library(lubridate)
library(tidyr)
library(dplyr)


data_long <- data %>%
  pivot_longer(cols = JAN:DEC,
               names_to = "Month",
               values_to = "Precipitation") %>%
    select(-YEAR.S.)
data_long$Month <- match(data_long$Month, toupper(month.abb))

data_long <- data_long %>%
  filter(!is.na(Precipitation))
head(data_long)

#Sample precipitation for given month

sample_monthly_precipitation <- function(month, n_samples, data_long) {
  historical_values <- data_long$Precipitation[data_long$Month== month]
  #Gen dens distribution
  dens <- density(historical_values)
  
  sampled_values <- sample(dens$x, size = n_samples, prob = dens$y, replace = TRUE)
  return(sampled_values)
}

#monte carlo simulation

rainfall_simulation <- function(num_trials, data_long) {
  count_exceed <- 0
  
  for (trial in 1:num_trials) {
    #simulate for each month
    simulated_annual_rainfall <- sapply(1:12, function(month) {
      mean(sample_monthly_precipitation(month, 1, data_long))
    })
    annual_total <- sum(simulated_annual_rainfall)
#check if any month > .25(annual_total)
    if (max(simulated_annual_rainfall) > 0.25 * annual_total) {
      count_exceed <- count_exceed +1
    }
  }
#return prob
  return(count_exceed/num_trials)
}

#run

num_trials <- 1000
result <- rainfall_simulation(num_trials, data_long)
print(paste("Probability that more than 25% of annual rainfall comes from a single month:", result))
```
```{r} 
install.packages("usethis", repos = "https://cloud.r-project.org")

```
