Monte Carlo Simulations reference 1/3

```{r}
library(rvest)
library(stringr)
library(readr)
library(tidyr)
```

Scraping from polymarket...

```{r}
get_R_win_prob = function(state) {
  # set base url for polymarket and get current unix time
  base_url = "https://polymarket.com/event/"
  current_unix_time <- as.integer(as.POSIXct(Sys.time(), origin="1970-01-01"))
  
  # read html page and extract Republican win probability
  # delaware, new york, and new mexico have different url patterns and require a different regex
  if(state %in% c("delaware", "new-york", "new-mexico")) {
    url = paste0(base_url, state, "-presidential-election-winner-2024/will-a-democrat-win-", state, "-presidential-election")
    pattern1 = "Democratic candidate wins.{0,1000}outcomePrices\\\":\\[\\\"0.[:digit:]*\\\""
    pattern2 = "outcomePrices\\\":\\[\\\"0.[:digit:]*\\\""
    page = read_html(url)
    text = html_text2(page)
    R_win_prob = 100-parse_number(str_extract(str_extract(text, pattern1), pattern2))*100
  } else {
    url = paste0(base_url, state, "-presidential-election-winner/will-a-republican-win-", state, "-in-the-2024-us-presidential-election?tid=", current_unix_time)
    pattern1 = "Republican candidate wins.{0,1000}outcomePrices\\\":\\[\\\"0.[:digit:]*\\\""
    pattern2 = "outcomePrices\\\":\\[\\\"0.[:digit:]*\\\""
    page = read_html(url)
    text = html_text2(page)
    R_win_prob = parse_number(str_extract(str_extract(text, pattern1), pattern2))*100
  }
  print(paste0("finished: ", state))
  return(R_win_prob)
}

# get list of states, change to lowercase, and change spaces to hyphens
states = str_to_lower(str_replace(state.name, " ", "-"))

# get Republican win probability for each state
R_win_probs = lapply(states, get_R_win_prob)

# create data frame with state names and win probability for each party
df = data.frame(State = state.name, R_win_prob = unlist(R_win_probs), D_win_prob = 100-(unlist(R_win_probs)))
```

```{r}
# get DC win probability

# get current unix time and set url
current_unix_time <- as.integer(as.POSIXct(Sys.time(), origin="1970-01-01"))
url = paste0("https://polymarket.com/event/washington-dc-presidential-election-winner/will-a-republican-win-washington-dc-presidential-election?tid=", current_unix_time)

# read html page
page = read_html(url)
text = html_text2(page)

pattern1 = "Democratic candidate wins.{0,1000}outcomePrices\\\":\\[\\\"0.[:digit:]*\\\""
pattern2 = "outcomePrices\\\":\\[\\\"0.[:digit:]*\\\""

R_win_prob_DC = 100-parse_number(str_extract(str_extract(text, pattern1), pattern2))*100

df_DC = data.frame(State = "District of Columbia", R_win_prob = R_win_prob_DC, D_win_prob = 100-(R_win_prob_DC))
df = rbind(df, df_DC)
df = df[order(df$State),]
```

```{r}
# get electoral vote counts
url = "https://www.archives.gov/electoral-college/allocation"
page = read_html(url)

# get table
electoral_table = data.frame(col = unname(unlist(html_table(html_nodes(page, "table")[[1]]))))

# separate column into state and electoral vote counts
electoral_table = separate(electoral_table, col, into = c("State", "Electoral_votes"), sep = " -")

# remove any non-numeric characters from electoral votes column
electoral_table$Electoral_votes = parse_number(electoral_table$Electoral_votes)
```

```{r}
# join electoral_table with df
df = merge(df, electoral_table)

# save table (just in case...)
#write.csv(df, "Polymarket_data.csv")
```

Basic Monte Carlo...

```{r}
# define function to predict electoral votes won by R for a given state
# v is a vector, one row from df
get_R_ev = function(v) {
  # sample 1 or 0 based on polymarket probabilities
  R_win = sample(c(1,0), 1, prob=c(v[2], v[3]))
  
  # get number of electoral votes earned by R
  R_ev = R_win*as.numeric(v[4])
  return(R_ev)
}

# set up Monte Carlo
predict_election = function(numTrials) {
  R_win_count = 0
  tie_count = 0
  
  for (trial in 1:numTrials) {
    # get predicted electoral votes for R for each state and find sum
    R_ev = sum(apply(df, 1, get_R_ev))
    # aggregate based on overall winner
    if(R_ev>=270) {
      R_win_count = R_win_count + 1
    } else if(R_ev==269) {
      tie_count = tie_count + 1
    }
  }
  # print results
  print(paste0("percentage ties: ", tie_count/numTrials*100, "%"))
  print(paste0("Percentage R won: ", R_win_count/numTrials*100, "%"))
  print(paste0("Percentage D won: ", (1-R_win_count/numTrials-tie_count/numTrials)*100, "%"))
  #return(R_win_count/numTrials)
}

# run Monte Carlo simulation for 1000 iterations
predict_election(1000)
```

Adding in some uncertainty...

```{r}
# v is a vector, one row from df
get_R_ev = function(v) {
  # mean win probability comes from polymarket
  R_win_prob_mean = as.numeric(v[2])
  # sample R_win_prob from normal distribution centered at polymarket probability
  # with standard deviation of 3%
  R_win_prob = rnorm(1, R_win_prob_mean, 3)
  # truncate results to the range 0 - 100
  if(R_win_prob>100) {R_win_prob = 100}
  if(R_win_prob<0) {R_win_prob = 0}
  # D win prob is complement of R win prob
  D_win_prob = 100-R_win_prob
  # sample 1 or 0 based on new probabilities
  R_win = sample(c(1,0), 1, prob=c(R_win_prob, D_win_prob))
  # calculate electoral votes
  R_ev = R_win*as.numeric(v[4])
  return(R_ev)
}

# Monte Carlo simulation...
predict_election = function(numTrials, mute_output = F) {
  R_win_count = 0
  tie_count = 0
  
  for (trial in 1:numTrials) {
    R_ev = sum(apply(df, 1, get_R_ev))
    if(R_ev>=270) {
      R_win_count = R_win_count + 1
    } else if(R_ev==269) {
      tie_count = tie_count + 1
    }
  }
  # new parameter (mute_output) allows you to hide the print statements
  if(!mute_output) {
    print(paste0("percentage ties: ", tie_count/numTrials*100, "%"))
    print(paste0("Percentage R won: ", R_win_count/numTrials*100, "%"))
    print(paste0("Percentage D won: ", (1-R_win_count/numTrials-tie_count/numTrials)*100, "%"))
  }
  # return data frame with aggregate simulation results
  return(data.frame(R_pct = R_win_count/numTrials, D_pct = 1-R_win_count/numTrials-tie_count/numTrials, tie_pct = tie_count/numTrials))
  #return(R_win_count/numTrials)
}

predict_election(1000)
```

```{r}
# get samples for plotting...
numTrials = 100
numRows = 1000
# create data frame with simulation results from 1000 simulations (numRows) of 100 trials (numTrials) each
tests = predict_election(numTrials)
for (i in 1:(numRows-1)) {
  tests = rbind(tests, predict_election(numTrials))
}
```

plot simulation results...

```{r}
# xdata is just simulation # (1 - numRows)
xdata = 1:nrow(tests)
# change margins to allow plotting legend outside main plot region
par(mar = c(5.1, 4.1, 4.1, 2.1), xpd=T)
# plot simulation results
plot(xdata, tests$R_pct, type='l', ylim = c(0,1), col='red', ylab="Probability", xlab="Sample #")
lines(xdata, tests$D_pct, col='blue')
lines(xdata, tests$tie_pct, col='black')
legend('topright', inset=c(-0.05, -0.15), legend=c(paste0('R win (avg = ', mean(tests$R_pct), ')'), paste0('D win (avg = ', mean(tests$D_pct), ')'), paste0('tie (avg = ', mean(tests$tie_pct), ')')), lty=1, col=c('red', 'blue', 'black'))
```

plot histogram of simulation results...

```{r}
tie_mean = mean(tests$tie_pct)
D_mean = mean(tests$D_pct)
R_mean = mean(tests$R_pct)

break_int = 0.02
# save histograms as objects so we can plot multiple together
tie_hist = hist(tests$tie_pct, breaks=seq(0,1,break_int), plot=F)
D_hist = hist(tests$D_pct, breaks=seq(0,1,break_int), plot=F)
R_hist = hist(tests$R_pct, breaks=seq(0,1,break_int), plot=F)

# plot histograms and change plotting parameters
plot(tie_hist, col=rgb(0.2,0.2,0.2,0.3), xlim=c(0,1), main = "Distribution of Simulation Results", xlab = 'Win probability')
# add vertical line for mean value
abline(v=tie_mean, col=rgb(0.2,0.2,0.2), lwd=2)
# add text for the value of mean tie percent
text(tie_mean+0.11, 400, paste0("mean = ", round(tie_mean, 3)))

plot(D_hist, col=rgb(0,0,1,0.3), xlim=c(0,1), add=T)
abline(v=mean(tests$D_pct), col='blue', lwd=2)
text(D_mean+0.1, 400, paste0("mean = ", round(D_mean, 3)))

plot(R_hist, col=rgb(1,0,0,0.3), xlim=c(0,1), add=T)
abline(v=R_mean, col='red', lwd=2)
text(R_mean+0.1, 400, paste0("mean = ", round(R_mean, 3)))

legend('topright', legend=c("Ties", "Democrat", "Republican"), fill=c(rgb(0.2,0.2,0.2,0.3), rgb(0,0,1,0.3), rgb(1,0,0,0.3)))
```

Variable uncertainty...

```{r}
# adding column for standard deviation
# [0-10: 0.5; 10-20: 1; 20-30: 2; 30-40: 3; 40-50:4]
df$sd = 0.5
df$sd[df$R_win_prob>10 & df$R_win_prob<90] = 1
df$sd[df$R_win_prob>20 & df$R_win_prob<80] = 2
df$sd[df$R_win_prob>30 & df$R_win_prob<70] = 3
df$sd[df$R_win_prob>40 & df$R_win_prob<60] = 4

# v is a vector, one row from df
get_R_ev = function(v) {
  R_win_prob_mean = as.numeric(v[2])
  # now, instead of sampling from normal distribution centered around polymarket value and standard deviation of 3, we will sample from normal distribution centered around polymarket value, but with standard deviation as set above (column 5 in df)
  R_win_prob = rnorm(1, R_win_prob_mean, as.numeric(v[5]))
  if(R_win_prob>100) {R_win_prob = 100}
  if(R_win_prob<0) {R_win_prob = 0}
  D_win_prob = 100-R_win_prob
  R_win = sample(c(1,0), 1, prob=c(R_win_prob, D_win_prob))
  R_ev = R_win*as.numeric(v[4])
  return(R_ev)
}

predict_election = function(numTrials, mute_output = F) {
  R_win_count = 0
  tie_count = 0
  
  for (trial in 1:numTrials) {
    R_ev = sum(apply(df, 1, get_R_ev))
    if(R_ev>=270) {
      R_win_count = R_win_count + 1
    } else if(R_ev==269) {
      tie_count = tie_count + 1
    }
  }
  if(!mute_output) {
    print(paste0("percentage ties: ", tie_count/numTrials*100, "%"))
    print(paste0("Percentage R won: ", R_win_count/numTrials*100, "%"))
    print(paste0("Percentage D won: ", (1-R_win_count/numTrials-tie_count/numTrials)*100, "%"))
  }
  return(data.frame(R_pct = R_win_count/numTrials, D_pct = 1-R_win_count/numTrials-tie_count/numTrials, tie_pct = tie_count/numTrials))
  #return(R_win_count/numTrials)
}

predict_election(1000)
```

Does our uncertainty really impact our results?

```{r}
# change functions so we can run simulations with different values of standard deviation (same value for all states in each simulation, different values for different simulations)
# v is a vector, one row from df
get_R_ev = function(v, sd) {
  R_win_prob_mean = as.numeric(v[2])
  R_win_prob = rnorm(1, R_win_prob_mean, sd)
  if(R_win_prob>100) {R_win_prob = 100}
  if(R_win_prob<0) {R_win_prob = 0}
  D_win_prob = 100-R_win_prob
  R_win = sample(c(1,0), 1, prob=c(R_win_prob, D_win_prob))
  R_ev = R_win*as.numeric(v[4])
  return(R_ev)
}

predict_election = function(numTrials, sd, mute_output = F) {
  R_win_count = 0
  tie_count = 0
  
  for (trial in 1:numTrials) {
    R_ev = sum(apply(df, 1, function(x) {get_R_ev(x, sd)}))
    if(R_ev>=270) {
      R_win_count = R_win_count + 1
    } else if(R_ev==269) {
      tie_count = tie_count + 1
    }
  }
  if(!mute_output) {
    print(paste0("percentage ties: ", tie_count/numTrials*100, "%"))
    print(paste0("Percentage R won: ", R_win_count/numTrials*100, "%"))
    print(paste0("Percentage D won: ", (1-R_win_count/numTrials-tie_count/numTrials)*100, "%"))
  }
  return(data.frame(R_pct = R_win_count/numTrials, D_pct = 1-R_win_count/numTrials-tie_count/numTrials, tie_pct = tie_count/numTrials))
}

```

```{r}
# get samples for plotting...
# we will try integer values of standard deviation from 0 through 80
x_data = seq(0,80,1)

numTrials = 1000
numRows = length(x_data)
tests = predict_election(numTrials, x_data[1], mute_output = T)
for (i in 2:numRows) {
  tests = rbind(tests, predict_election(numTrials, x_data[i], mute_output = T))
}

```

```{r}
# make plot with standard deviation value on x-axis and win probabilities on y-axis
plot(x_data, tests$R_pct, type='l', ylim = c(0,1), col='red', ylab="Probability", xlab="sd")
lines(x_data, tests$D_pct, col='blue')
lines(x_data, tests$tie_pct, col='black')
legend('topright', legend=c('Republican', 'Democrat', 'Tie'), lty=1, col=c('red', 'blue', 'black'))
```

What if we wanted to join our data frame to a shapefile of state outlines and put it on arcgis online?

```{r}
require(sf)
# find shapefile online and unzip
# dsn is the folder with your shapefile, layer is the base name of your shapefile (i.e. "filename.shp" minus ".shp")
shape <- read_sf(dsn = "U:/Documents/econ8600/10-29-24/cb_2018_us_state_500k", layer = "cb_2018_us_state_500k")

# check to make sure the columns match, we want 51 rows (1 for each state and 1 for DC)
length(intersect(df$State, shape$NAME))

# make column names match
names(shape)[names(shape)=="NAME"]="State"

# merge the tables (inner join)
out_sf = merge(shape, df)

# save output as a shapefile
st_write(out_sf, "election_probabilities.shp")

# next steps are outside of R...
# compress your new files to a zipfile
# upload to ArcGIS online and create layer
# change layer symbology and create map that you can use in dashboard

# this would work for other census boundaries or countries AS LONG AS there is a "key" column, a column with matching values that you can join by (could be country name or some ID value)

```

We can visualize this in R as well... (ChatGPT used for generating maps, I don't do ggplot...)

These visualizations are rough, but quick. I wouldn't recommend trying to make "publication-ready" maps in R, but you can use it to check that your data looks right and quickly look for spatial patterns.

```{r}
library(ggplot2)
library(plotly)

ggplot(data = out_sf) +
  geom_sf(aes(fill = R_win_prob)) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 50) +
  coord_sf(xlim = c(-180, -60), ylim = c(18, 72)) + 
  theme_minimal() +
  labs(title = "Republican Winning Probability by State",
       fill = "R Win Probability") +
  theme(panel.grid = element_blank())

```

interactive map...

```{r}

p <- ggplot(data = out_sf) +
  geom_sf(aes(fill = R_win_prob, text = paste("State:", State, "<br>R Win Prob:", round(R_win_prob, 2), "<br>D Win Prob:", round(D_win_prob,2)))) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 50) +
  coord_sf(xlim = c(-180, -60), ylim = c(18, 72)) + 
  theme_minimal() +
  labs(title = "Republican Winning Probability by State",
       fill = "R Win Probability") +
  theme(panel.grid = element_blank())

# Convert ggplot object to plotly for interactivity
p_interactive <- ggplotly(p, tooltip = "text")  # Specify tooltip to show

# Display the interactive map
p_interactive

```
