## Random Forest

Using Boston housing data from [https://www.cs.toronto.edu/\~delve/data/boston/bostonDetail.html](#0){.uri}

Each record in the dataset describes a Boston suburb or town. The data was drawn from the Boston Standard Metropolitan Statistical Area (SMSA) in 1970.

### Step 1: Load the Dataset

Begin by loading the Boston Housing dataset from the `MASS` package.

```{r}
# Install MASS package if not already installed
install.packages("MASS")

# Load the required package
library(MASS)

# Look at the details of the dataset
?Boston

# Load the dataset
data = Boston

# View the summary of the dataset
summary(data)

```

### Step 2: Explore Dataset Characteristics

Generate boxplots for each variable

```{r}
library(RColorBrewer)
colors = colorRampPalette(brewer.pal(12, "Set3"))(length(names(data)))  # Expand Set3 palette to match the number of variables

par(mfrow = c(3, 5), mar = c(1, 2, 2, 1))  # Arrange plots in a 3x5 grid to fit all variables
# Loop through each variable, plotting with a unique color
for (i in seq_along(names(data))) {
  boxplot(data[[i]], main = names(data)[i], col = colors[i], border = "darkgray", ylab = "")
}

# Reset plotting layout
par(mfrow = c(1, 1))
```

Create histograms with kernel density for each variable

```{r}
# Set up a grid layout with adjusted margins
par(mfrow = c(3, 5), mar = c(2, 2, 2, 1))  # Adjust layout as needed

# Create histograms for each variable
lapply(names(data), function(var) {
  hist(data[[var]], main = var, xlab = "", ylab = "", col = "lightblue", border = "white", probability = TRUE)  # Set probability = TRUE to scale histogram for density
  lines(density(data[[var]]), col = "red", lwd = 2)  # Overlay density line
})

# Reset layout
par(mfrow = c(1, 1))
```

Plot the correlation heatmap

```{r}
#install.packages("corrplot")
library(corrplot)

# Calculate correlation matrix
correlation_matrix = cor(Boston)

# Plot the correlation heatmap
# Play around with different options for the method parameter...
corrplot(correlation_matrix, method = "ellipse" #cikir =, type = "upper", 
         tl.col = "black", tl.srt = 45, addCoef.col = "black", number.cex = 0.6,
         col = colorRampPalette(c("blue", "white", "red"))(200))

```

### Step 2: Install and Load Required Packages for random forest

```{r}
# Install randomForest package if not already installed
#install.packages("randomForest")

# Load the randomForest package
library(randomForest)
#randomly bags
```

### 

### Step 4: Train the Random Forest Model

Train the random forest model to predict `medv` based on all the other features.

```{r}
# Train the model
set.seed(123)  # for reproducibility
rf_model = randomForest(medv ~ ., data = data, ntree = 500, importance = TRUE)
#can also omit ntree, and it will decide for you and that's the number of decision trees
#importance = TRUE means i am interested in looking at variable importance after run. Random Forest tells you how important each variable is
# View model summary
print(rf_model)

```

### Step 5: Interpret Feature Importance

Random forest provides a measure of feature importance, showing how much each variable contributes to predictions.

-   Which variables appear to have the highest importance?

-   How could variables like `lstat` (percent lower status of the population) or `rm` (average number of rooms) intuitively impact housing prices?

```{r}
# Extract and plot feature importance
importance = importance(rf_model)
varImpPlot(rf_model)

```

### 1. **%IncMSE (Percent Increase in Mean Squared Error)**

-   **What It Means**: `%IncMSE` represents the increase in prediction error (Mean Squared Error) when the values of a specific predictor are randomly permuted while all other variables remain the same. This method is often called "permutation importance."

-   **How It Works**: For each feature, the algorithm shuffles the values in that column, making the feature less informative. It then recalculates the model's prediction error with this "broken" feature.

-   **Interpretation**: If the `%IncMSE` for a feature is high, it means that the feature is important for making accurate predictions—its absence or scrambling has a significant negative impact on the model’s accuracy. Conversely, a low `%IncMSE` suggests the feature has minimal impact on the model’s predictive power.

### 2. **IncNodePurity (Increase in Node Purity)**

-   **What It Means**: `IncNodePurity` measures the total decrease in node impurities across all trees in the forest due to each feature. For regression, node impurity is often calculated using variance reduction, while for classification, it might be based on Gini impurity or entropy.

-   **How It Works**: Every time a feature is used to split a node, it reduces the impurity (or disorder) within that node by creating a more homogeneous group. `IncNodePurity` aggregates the impurity reductions each feature contributes across all trees in the random forest.

-   **Interpretation**: Features with a higher `IncNodePurity` have a stronger influence on creating splits that reduce variance in the target variable. A high `IncNodePurity` indicates that the feature plays a significant role in improving the model’s structure, while a low value means it’s less useful in splitting nodes effectively.

### Which Metric to Focus On

-   **For Interpretability**: `%IncMSE` is often more interpretable, as it directly relates to how much the feature influences prediction accuracy.

-   **For Structure**: `IncNodePurity` can offer insight into the structural importance of features within the trees but might be less directly tied to prediction accuracy.

### Step 6: Generate Predictions and Evaluate Model Performance

Have students generate predictions and evaluate model performance, using mean squared error (MSE) as an example metric.

```{r}
# Generate predictions
predictions = predict(rf_model, data)

# Calculate Mean Squared Error
mse = mean((data$medv - predictions)^2)
print(paste("Mean Squared Error:", round(mse, 2)))

```
#out of bag - leave one out, model training
```{r}
# Generate predictions using the random forest model
predictions = predict(rf_model, Boston)
observed = Boston$medv  # Actual observed values

# Create the plot
plot(observed, predictions, 
     main = "Predicted vs. Observed Housing Prices",
     xlab = "Observed MEDV", 
     ylab = "Predicted MEDV", 
     pch = 16, 
     col = rgb(0.2, 0.4, 0.6, 0.7))

# Add a reference line for a perfect fit
abline(0, 1, col = "red", lwd = 2, lty = 2)

```

### Step 7: Use Partial Dependence Plots

If you want to dive deeper, partial dependence plots (PDP) can illustrate how certain variables affect predictions. Here’s an example with the `pdp` package:

```{r}
# Install and load pdp package
#install.packages("pdp")
library(pdp)

# Create a partial dependence plot
var = "nox"
pdp_lstat = partial(rf_model, pred.var = var)
plotPartial(pdp_lstat, lwd = 2, main = paste0("Partial Dependence of MEDV on ", var))

```

### Discussion Points

1.  **Feature Importance**: Talk through which features are most influential and why.

2.  **Predictions**: Compare predicted vs. actual `medv` values to discuss model accuracy.

3.  **Partial Dependence**: For selected features, discuss the relationship shown in PDPs and connect it back to real-world understanding.
