## Random Forest (updated version)

Using Boston housing data from [https://www.cs.toronto.edu/\~delve/data/boston/bostonDetail.html](#0){.uri}

Each record in the dataset describes a Boston suburb or town. The data was drawn from the Boston Standard Metropolitan Statistical Area (SMSA) in 1970.

500 is pretty small for a random forest experiment.

### Step 1: Load the Dataset

Begin by loading the Boston Housing dataset from the `MASS` package.

```{r}
# Install MASS package if not already installed

# Load the required package
library(MASS)

# Look at the details of the dataset
?Boston

# Load the dataset
data = Boston

# View the summary of the dataset
summary(data)

```

### Step 2: Explore Dataset Characteristics

Generate boxplots for each variable

```{r}
library(RColorBrewer)
colors = colorRampPalette(brewer.pal(12, "Set3"))(length(names(data)))  # Expand Set3 palette to match the number of variables

par(mfrow = c(3, 5), mar = c(1, 2, 2, 1))  # Arrange plots in a 3x5 grid to fit all variables
# Loop through each variable, plotting with a unique color
for (i in seq_along(names(data))) {
  boxplot(data[[i]], main = names(data)[i], col = colors[i], border = "darkgray", ylab = "")
}

# Reset plotting layout
par(mfrow = c(1, 1))
```

Create histograms with kernel density for each variable

```{r}
# Set up a grid layout with adjusted margins
par(mfrow = c(3, 5), mar = c(2, 2, 2, 1))  # Adjust layout as needed

# Create histograms for each variable
lapply(names(data), function(var) {
  hist(data[[var]], main = var, xlab = "", ylab = "", col = "lightblue", border = "white", probability = TRUE)  # Set probability = TRUE to scale histogram for density
  lines(density(data[[var]]), col = "red", lwd = 2)  # Overlay density line
})

# Reset layout
par(mfrow = c(1, 1))
```

Plot explanatory vs. response

```{r}
#isn't there an easier way to do a correlation matrix of the variables
# Set up a grid layout with adjusted margins
par(mfrow = c(3, 5), mar = c(2, 2, 2, 1))  # Adjust layout as needed

# Create histograms for each variable
lapply(names(data), function(var) {
  plot(data[[var]], data$medv, pch=16, col=rgb(0, 0, 0, 0.1), main = var, xlab = "", ylab = "") #opacity shows the density best
})

# Reset layout
par(mfrow = c(1, 1))
```

Plot the correlation heatmap

```{r}
#install.packages("corrplot")
library(corrplot)

# Calculate correlation matrix
correlation_matrix = cor(Boston)

# Plot the correlation heatmap
# Play around with different options for the method parameter...
corrplot(correlation_matrix, method = "color", type = "upper", 
         tl.col = "black", tl.srt = 45, addCoef.col = "black", number.cex = 0.6,
         col = colorRampPalette(c("blue", "white", "red"))(200))

```
for example, above the rm variable suggests that home values increase with distance from employment centers.

### Step 2: Install and Load Required Packages for random forest

```{r}
# Install randomForest package if not already installed
#install.packages("randomForest")

# Load the randomForest package
library(randomForest)

```

### 

### Step 4: Train the Random Forest Model

Train the random forest model to predict `medv` based on all the other features.

```{r}
# Train the model
set.seed(123)  # for reproducibility
rf_model = randomForest(medv ~ ., data = data, ntree = 500, importance = TRUE)
  #medv ~. is the thing that says all the variables, no interactions? 
# View model summary
print(rf_model)

```

### Step 5: Interpret Feature Importance

Random forest provides a measure of feature importance, showing how much each variable contributes to predictions.
-   purity - goes into a node, splits based on some criteria like #rooms= 2 

-   Which variables appear to have the highest importance?

-   How could variables like `lstat` (percent lower status of the population) or `rm` (average number of rooms) intuitively impact housing prices?

```{r}
# Extract and plot feature importance
importance = importance(rf_model)
varImpPlot(rf_model)

```

### 1. **%IncMSE (Percent Increase in Mean Squared Error)**

-   **What It Means**: `%IncMSE` represents the increase in prediction error (Mean Squared Error) when the values of a specific predictor are randomly permuted while all other variables remain the same. This method is often called "permutation importance."
*HOW DOES INC ERROR OF ONE PREDICTOR IMPACT MSE of model?*

-   **How It Works**: For each feature, the algorithm shuffles the values in that column, making the feature less informative. It then recalculates the model's prediction error with this "broken" feature.

-   **Interpretation**: If the `%IncMSE` for a feature is high, it means that the feature is important for making accurate predictions—its absence or scrambling has a significant negative impact on the model’s accuracy. Conversely, a low `%IncMSE` suggests the feature has minimal impact on the model’s predictive power.
  - number of rooms and lstat, highly positively correlated and see
    - have a multiple linear model, and one same marginal inc in predictor. 10-12 same change in lstat as in 
    - rf doesn't assume constant relationship
      - partial dependence- 

### 2. **IncNodePurity (Increase in Node Purity)**

-   **What It Means**: `IncNodePurity` measures the total decrease in node impurities across all trees in the forest due to each feature. For regression, node impurity is often calculated using variance reduction, while for classification, it might be based on Gini impurity or entropy.

-   **How It Works**: Every time a feature is used to split a node, it reduces the impurity (or disorder) within that node by creating a more homogeneous group. `IncNodePurity` aggregates the impurity reductions each feature contributes across all trees in the random forest.

-   **Interpretation**: Features with a higher `IncNodePurity` have a stronger influence on creating splits that reduce variance in the target variable. A high `IncNodePurity` indicates that the feature plays a significant role in improving the model’s structure, while a low value means it’s less useful in splitting nodes effectively.

### Which Metric to Focus On

-   **For Interpretability**: `%IncMSE` is often more interpretable, as it directly relates to how much the feature influences prediction accuracy.

-   **For Structure**: `IncNodePurity` can offer insight into the structural importance of features within the trees but might be less directly tied to prediction accuracy.

### Step 6: Generate Predictions and Evaluate Model Performance

Generate predictions and evaluate model performance, using mean squared error (MSE) as an example metric.

```{r}
# Generate predictions. Scatterplot with observed vs. predicted values
predictions = predict(rf_model, data) #trained on ENTIRE dataset, then applying to model that already seen, already built decision trees based on it.

# Calculate Mean Squared Error
mse = mean((data$medv - predictions)^2)
print(paste("Mean Squared Error:", round(mse, 2)))

```

```{r}
# Generate predictions using the random forest model
predictions = predict(rf_model, Boston)
observed = Boston$medv  # Actual observed values

# Create the plot
plot(observed, predictions, 
     main = "Predicted vs. Observed Housing Prices",
     xlab = "Observed MEDV", 
     ylab = "Predicted MEDV", 
     pch = 16, 
     col = rgb(0.2, 0.4, 0.6, 0.7))

# Add a reference line for a perfect fit
abline(0, 1, col = "red", lwd = 2, lty = 2)

```
Get's bad at 50 - artificially cut off at 50 so distribution drops off and then jumps up around 50 
  - data[data$medv>46,] #prints all rows and columns where > 46, only a couple in there that are not 50. nrow(data[data$medv>46,]) doesn't seem very likely, probably censored at 50, so issues with predictability. Exactly 50, model didn't predict any of them to be exactly 50 so prediction of higher values not as good.


### Step 7: Use Partial Dependence Plots

If you want to dive deeper, partial dependence plots (PDP) can illustrate how certain variables affect predictions. Here’s an example with the `pdp` package:

```{r}
# Install and load pdp package
#install.packages("pdp")
library(pdp)

# Create a partial dependence plot
var = "dis" # lstat, dis & rm  also looked at in class.
pdp_lstat = partial(rf_model, pred.var = var)
plotPartial(pdp_lstat, lwd = 2, main = paste0("Partial Dependence of MEDV on ", var), rug=T, train=data)
#the rug is what gives the tick marks to interpret pdp's further
#lstat has a lot more spread/data/variation of the predictor

```
super close, think maybe penthouses in NYC, then  decrease for poor public housing, then inc again for suburbs, and then decrease when ruralness not good.
- only 30% is actually positive/increasing.

Exponential change - s-curve shape, logarithmic? all of the effect captured between 6 & 8, the small tick marks on rm axis show percentiles. (dont have a ton of data below 5.7, or above 7.2.  80% of data falls between that middle range for the predictor rm

### Data Partitioning

If we are using a larger dataset, we would want to split our data into training and testing data to see how well our model performs with data it hasn't been trained on.

#some portion train, some portion test. How well does your model predict data it's never seen. 20-30% testing so 70-80% for training.

```{r}
## 75% of the sample size
smp_size <- floor(0.75 * nrow(data)) #floor tells to round down to nearest whole number. 

## set the seed to make your partition reproducible
set.seed(123)
train_ind <- sample(seq_len(nrow(data)), size = smp_size) 

train <- data[train_ind, ]
test <- data[-train_ind, ] #all that are NOT in train_ind

#dim(train) to make sure correct
#dim(test)
```

Then rerun our model...

```{r}
rf_model = randomForest(medv ~ ., data = train, ntree = 500, importance = TRUE)
#now tell only want to train on the 75%, using the random data partitioned for our 

# View model summary
print(rf_model)
```
split as in the split in the decision trees?

For our predictions, we'll use the testing data...

```{r}
# Generate predictions
predictions = predict(rf_model, test) 

# Calculate Mean Squared Error
mse = mean((test$medv - predictions)^2)
print(paste("Mean Squared Error:", round(mse, 2)))
```
MSE increased to 12.84! so a lot worse at predicting values it hasn't seen before.
And plot predicted vs. observed

```{r}
# Generate predictions using the random forest model
predictions = predict(rf_model, test)
observed = test$medv  # Actual observed values

# Create the plot
plot(observed, predictions, 
     main = "Predicted vs. Observed Housing Prices",
     xlab = "Observed MEDV", 
     ylab = "Predicted MEDV", 
     pch = 16, 
     col = rgb(0.2, 0.4, 0.6, 0.7))

# Add a reference line for a perfect fit
abline(0, 1, col = "red", lwd = 2, lty = 2)
```
See that in the plot above, a lot more spread from the line 
overpredicted even worse.OVERFIT model when give full dataset.

Partial dependence plots have a lot going on and are more difficult to interpret. 