
Starting with stock history scraping
```{r}
# Load required libraries
library(httr)
library(rvest)
library(stringr)
library(dplyr)

get_stock_history = function(ticker, start_time = 1) {
  # Convert to Unix time
  current_unix_time <- as.integer(as.POSIXct(Sys.time(), origin="1970-01-01"))
  
  base_url = "https://finance.yahoo.com/quote/"
  url = paste0(base_url, ticker, "/history/?period1=1&period2=", current_unix_time)
  
  # Set user agent and make the GET request
  response <- GET(url, add_headers("User-Agent" = "Mozilla/5.0 (Windows NT 10.0; Win64; x64) AppleWebKit/537.36 (KHTML, like Gecko) Chrome/91.0.4472.124 Safari/537.36",
    "Accept" = "text/html,application/xhtml+xml,application/xml;q=0.9,image/webp,*/*;q=0.8",
    "Referer" = "https://finance.yahoo.com/",
    "Accept-Language" = "en-US,en;q=0.5",
    "Connection" = "keep-alive"))
  
  # Check response status
  if (http_status(response)$category == "Success") {
      # Read the content of the response
      page <- content(response, "text")
      
      # Parse the HTML content
      parsed_page <- read_html(page)
  
      # Check the structure of the parsed page to find the correct table
      # You can print the HTML content to debug if needed
      # print(parsed_page)
  
      # Attempt to extract the historical prices table
      data_table <- html_table(html_nodes(parsed_page, "table")[[1]])
  
      # Check if the table was extracted
      if (!is.null(data_table)) {
          # Clean the table (if necessary)
          cleaned_data <- rename(data_table,
                                 "Close" = "Close          Close price adjusted for splits.",
                                 'Adj_close' = 'Adj Close          Adjusted close price adjusted for splits and dividend and/or capital gain distributions.')
          cleaned_data = mutate(cleaned_data, Date = as.Date(Date, format = "%b %d, %Y"))
          cleaned_data$Volume = str_replace_all(cleaned_data$Volume, ",", "")
          cleaned_data[,-1] = apply(cleaned_data[,-1], 2, as.numeric)
  
          # Return the cleaned data
          return(cleaned_data)
      } else {
          print("Table extraction failed. No data found.")
      }
  } else {
      print(http_status(response))
  }
}

```

```{r}
# get stock history for Microsoft...
MSFT = na.omit(get_stock_history('MSFT'))

# calculate daily change (absolute and percent)
MSFT$Daily_change = MSFT$Close - MSFT$Open
MSFT$Daily_pct_change = MSFT$Daily_change/MSFT$Open

# plot closing price over time
plot(MSFT$Date, MSFT$Close, type='l')
```


Monte Carlo example from previous lecture

```{r}
# random walk to calculate theoretical stock price after ndays
# 1. Pick a space of possible samples (domain).
random_walk = function(start_price, days, mean, sd) {
  price = start_price * (1 + rnorm(1, mean, sd))
  for (i in 1:(days)) {
    price = price * (1  + rnorm(1, mean, sd))
  }
  return(price)
}

stock_monte_carlo = function(numTrials, starting_price, ndays, mean, sd) {
  count_increase = 0
  for (i in 1:numTrials) {
    # 2. Randomly sample in that space using a probability distribution.
    new_stock_price = random_walk(starting_price, ndays, mean, sd)
    
    # 3. Perform defined operations on the random samples.
    if (new_stock_price>starting_price) {
      # 4. Aggregate the data.
      count_increase = count_increase + 1
    }
  }
  # return percent of trials that meet our criteria (new_stock_price>starting_price)
  return(count_increase/numTrials)
}

# set parameter values
starting_price = 100
ndays = 30
mean = 0.001
sd = 0.005

# run simulation 1,000 times...
stock_monte_carlo(1000, starting_price, ndays, mean, sd)
```

We'll need to change the probability distribution we're sampling from...

We can look at a normal distribution plotted over the histogram of daily percent changes:

```{r}
# Generate your data
x_data = seq(-0.1, 0.1, by = 0.001)

# Create a histogram
hist(MSFT$Daily_pct_change, 
     freq = FALSE, 
     breaks = 100,  # number of breaks
     xlim = range(x_data),
     col = "lightblue",  # Histogram fill color
     border = "black",    # Border color for the histogram
     main = "Histogram of Daily Percentage Change",  # Main title
     xlab = "Daily Percentage Change",  # X-axis label
     ylab = "Density")  # Y-axis label

# Add normal distribution line
dnorm_values = dnorm(x_data, mean = mean(MSFT$Daily_pct_change), sd = sd(MSFT$Daily_pct_change))
lines(x_data, dnorm_values, col = "red", lwd = 2)

```

That's pretty close, but it doesn't match perfectly.
What if we fudge the standard deviation?
```{r}
# Create a histogram
hist(MSFT$Daily_pct_change, 
     freq = FALSE, 
     breaks = 100,  # number of breaks
     xlim = range(x_data),
     col = "lightblue",  # Histogram fill color
     border = "black",    # Border color for the histogram
     main = "Histogram of Daily Percentage Change",  # Main title
     xlab = "Daily Percentage Change",  # X-axis label
     ylab = "Density")  # Y-axis label

# Add normal distribution line
dnorm_values = dnorm(x_data, mean = mean(MSFT$Daily_pct_change), sd = 0.014)
lines(x_data, dnorm_values, col = "red", lwd = 2)
```

Still not perfect and we're manipulating the data a bit if we use a distribution like this.

Thankfully, we can plot the probability distribution of our data!
```{r}
dens = density(MSFT$Daily_pct_change)

# Create a histogram
hist(MSFT$Daily_pct_change, 
     freq = FALSE, 
     breaks = 100,  # number of breaks -- What are breaks?
     xlim = range(x_data),
     col = "lightblue",  # Histogram fill color
     border = "black",    # Border color for the histogram
     main = "Density Plot of Daily Percentage Change",  # Main title
     xlab = "Daily Percentage Change",  # X-axis label
     ylab = "Density")  # Y-axis label

# plot density function
lines(dens, col='red', lwd=2)

```

And we can sample values directly from the probability distribution of our data
```{r}
# Draw random values based on the density
# Use the density values as weights
n_samples <- 100  # Specify how many samples you want to draw
sampled_values <- sample(dens$x, size = n_samples, prob = dens$y, replace = TRUE)

# Plot the density and the sampled values
plot(dens, main = "Density Plot with Sampled Values", xlim = range(x_data), xlab = "Value", ylab = "Density", col = "red")
points(sampled_values, rep(0, length(sampled_values)), col = rgb(0,0,1,.1), pch = 16, cex = 1)  # Overlay sampled values

```

Now we can update the probability distribution from which we're sampling for our random walk simulation
```{r}
# random walk to calculate theoretical stock price after ndays
# 1. Pick a space of possible samples (domain).
random_walk = function(start_price, ndays, dens) {
  price = start_price * (1 + sample(dens$x, 1, prob = dens$y))
  for (i in 1:(ndays)) {
    price = price * (1  + sample(dens$x, 1, prob = dens$y))
  }
  return(price)
}

stock_monte_carlo = function(numTrials, starting_price, ndays, density) {
  count_increase = 0
  for (i in 1:numTrials) {
    # 2. Randomly sample in that space using a probability distribution.
    new_stock_price = random_walk(starting_price, ndays, density)
    
    # 3. Perform defined operations on the random samples.
    if (new_stock_price>starting_price) {
      # 4. Aggregate the data.
      count_increase = count_increase + 1
    }
  }
  # return percent of trials that meet our criteria (new_stock_price>starting_price)
  return(count_increase/numTrials)
}

# set parameter values
starting_price = 100
ndays = 30
density = density(MSFT$Daily_pct_change)

# run simulation 1,000 times...
stock_monte_carlo(1000, starting_price, ndays, density)
```

plot results against ndays

```{r}
ndays = 1:100
numTrials = 100
sim_results = unlist(lapply(ndays, function(x){stock_monte_carlo(numTrials, starting_price, x, density)}))

plot(ndays, sim_results, 
     type = 'l', 
     col = "black", 
     lwd = 2, 
     main = "Monte Carlo Simulation of Stock Prices",
     xlab = "Days", 
     ylab = "Probability of price increase", 
     #ylim = range(sim_results, na.rm = TRUE),  # Adjust y-axis limits based on data
     ylim = c(0,1),
     las = 1)  # Set axis labels to be horizontal

# add trendline
trend = lm(sim_results ~ ndays)$coefficients
abline(trend[1], trend[2], col="blue", lwd=2)

# add coin flip line...
abline(.5,0,col='red', lwd=2)

# add legend
legend('bottomright', legend=c("Simulation results", paste0("Trend (", round(trend[2],6), ")"), "50/50 line"), lty=1, lwd=2, col=c("black", "blue", "red"))
```


Simulate based on all data from the month of November (in any year)
```{r}
library(lubridate)

stock_monte_carlo(1000, starting_price, ndays=30, density(MSFT$Daily_pct_change[month(MSFT$Date) == 11]))

```


Plot simulation results for each calendar month
```{r}
numTrials = 200 # number of Monte Carlo trials for each month
monthly_results = unlist(lapply(1:12, function(x) {stock_monte_carlo(numTrials, starting_price, ndays=30, density(MSFT$Daily_pct_change[month(MSFT$Date) == x]))}))


plot(1:12, monthly_results, type='l')
```


generate 100 sets of monthly results for 100 iterations each of our monte carlo function
```{r}
nrows = 100
numTrials = 100
ndays = 30

monthly_results = unlist(lapply(1:12, function(x) {stock_monte_carlo(numTrials, starting_price, ndays, density(MSFT$Daily_pct_change[month(MSFT$Date) == x]))}))

for (i in 2:nrows) {
  monthly_results = rbind(monthly_results, unlist(lapply(1:12, function(x) {stock_monte_carlo(numTrials, starting_price, ndays, density(MSFT$Daily_pct_change[month(MSFT$Date) == x]))})))
}

matplot(1:12, t(monthly_results), type='l', col=rgb(1, 0, 0, 0.2))
```

```{r}
colnames(monthly_results) <- month.abb

boxplot(monthly_results, 
        main = "Boxplots of Monthly Data", 
        xlab = "Months", 
        ylab = "Values", 
        col = "lightblue", 
        border = "black")
abline(.5,0, lwd=2, col='darkred')
```
